---
title: "Lab 05: Introduction to Time Series Regressions - Part II"
subtitle: "Econometrics II - YTU"
author: 
  name: "Prof. Dr. Hüseyin Taştan"
  affiliation: "Yıldız Technical University"
# date: "`r format(Sys.time(), '%d %B %Y')`"
date: 2021 Spring
output: 
  html_document:
    number_sections: true
    theme: readable
    highlight: haddock 
    # code_folding: show
    toc: yes
    toc_depth: 3
    toc_float: yes
    keep_md: true
---
<style type="text/css"> 
body{
  font-size: 12pt;
}
code.r{
  font-size: 12pt;
}
</style>

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = TRUE, results = 'asis', fig.show = 'asis')
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
``` 

# Replicating Examples in ch.11 of Wooldridge's Text (6th ed) 

## Example 11.4: Efficient Market Hypothesis

Can we beat the market?

```{r, message=FALSE, warning=FALSE}
library(wooldridge)
library(dynlm)
library(stargazer)
data(nyse, package='wooldridge') 
nysets <- ts(nyse)
head(nysets)
```

Note that because return is defined as 
\[
return_t = 100\times \frac{price_t-price_{t-1}}{price_{t-1}}
\]
the first observation is NA (missing). Also, when we take the first lag of return we lose another observation (see the column `return_1`. The same applies to the changes as well (ie., cprice) which is defined as 
\[
cprice_t \equiv \Delta price_t = price_t - price_{t-1}
\]


Estimate AR(p) models using OLS and summarize the results in a table using `stargazer` package: 
```{r}
# Dynamic models: AR(p)
reg1 <- dynlm(return ~ L(return)                        , data=nysets) 
reg2 <- dynlm(return ~ L(return)+L(return,2)            , data=nysets) 
reg3 <- dynlm(return ~ L(return)+L(return,2)+L(return,3), data=nysets) 
# Regression Table
stargazer(reg1, reg2, reg3, type="text",
                            keep.stat=c("n","rsq","adj.rsq","f"))
```

**Exercise**: Download Google stock data and estimate AR models with lags 1, 2, and 3. 
**Solution**: Google stock market data data is available in `GOOGL.RData` file in Data folder. This was downloaded from Yahoo Finance using `quantmod` package. Here is the commands I used: 
```{r, eval=FALSE}
# Download daily stock market data using quantmod package
# and save the resulting data set (xts object)
library(quantmod) # need this to download stock market data from Yahoo Finance
# Download data using the quantmod package:
getSymbols("GOOGL", auto.assign = TRUE)
# save(GOOGL, file="Data/GOOGL.RData")
```

Load the data set and display the first six observations:
```{r}
library(forecast)
library(zoo)
library(dynlm)
library(stargazer)
library(ggplot2)
load("../Data/GOOGL.RData")
head(GOOGL)
```
```{r}
autoplot(GOOGL[, "GOOGL.Adjusted"]) +
  ggtitle("Google Daily Closing Prices") + xlab("")
```


```{r}
# Calculate return as the log difference
googret <- diff(log(GOOGL$GOOGL.Adjusted) )

# Plot returns
autoplot(googret) + xlab("")
```


```{r}
# Dynamic models: Ar(1), AR(2), AR(3)
googret <- as.zoo(googret)  # dynlm cannot handle xts objects
reg1 <- dynlm(googret~L(googret) ) 
reg2 <- dynlm(googret~L(googret)+L(googret,2) ) 
reg3 <- dynlm(googret~L(googret)+L(googret,2)+L(googret,3) ) 

# regression table
stargazer(reg1, reg2, reg3, type="text",
                            keep.stat=c("n","rsq","adj.rsq","f"))
```

**EXercise**: Can we predict Google returns? 

## Example 11.5: Expectations Augmented Phillips Curve

```{r}
library(forecast)
library(wooldridge) 
library(dynlm) 
tsphillips <- ts(phillips, start=1948, end=1996, frequency = 1)
reg1 <- dynlm(d(inf) ~ unem, data = tsphillips)
summary(reg1)
```


```{r}
# plot the residuals
uhat <- resid(reg1) 
autoplot(uhat)+ 
  geom_hline(yintercept = 0,linetype="dashed", color = "red")
```

## Example 11.6: Fertility Equation in First Differences


```{r}
library(dynlm)
library(stargazer)
data(fertil3, package='wooldridge')

# Define Yearly time series beginning in 1913
fertil_ts <- ts(fertil3, start=1913)
# display the first 5 autocorrelations of gfr
(acf(fertil_ts[, "gfr"], 5))
```

```{r}
# Regression in first differences:
res1 <- dynlm( d(gfr) ~ d(pe), data=fertil_ts)

# FDL in first differences
res2 <- dynlm( d(gfr) ~ d(pe) + L(d(pe)) + L(d(pe),2), data=fertil_ts)

# table
stargazer(res1, res2, type="text") 
```

## Example 11.7: Wages and Productivity


```{r}
# Regression with linear trend
library(dynlm)
library(wooldridge)
data(earns, package='wooldridge')

# Define ts object
earns_ts <- ts(earns, start=1947, end=1987)

fit1 <- dynlm( log(hrwage) ~ log(outphr) + trend(earns_ts), data=earns_ts)
summary(fit1)
```


```{r}
# Regression in first differences
fit2 <- dynlm( d(log(hrwage)) ~ d(log(outphr)), data=earns_ts)
summary(fit2)
```

**Exercise** Write down both models in equation form and interpret the results. 

<br>
<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>


