---
title: "Lab 09: Forecasting"
subtitle: "Econometrics II - YTU"
author: 
  name: "Prof. Dr. Hüseyin Taştan"
  affiliation: "Yıldız Technical University"
# date: "`r format(Sys.time(), '%d %B %Y')`"
date: 2021 Spring
output: 
  html_document:
    number_sections: true
    theme: readable
    highlight: haddock 
    # code_folding: show
    toc: yes
    toc_depth: 3
    toc_float: yes
    keep_md: true
---
<style type="text/css"> 
body{
  font-size: 12pt;
}
code.r{
  font-size: 12pt;
}
</style>

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = TRUE, results = 'asis', fig.show = 'asis')
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
``` 

# Basics of Forecasting  

## Example: Forecasting Unemployment 


Load packages, data set, and plot 3-month treasury bill rates: 
```{r}
library(dynlm) 
library(stargazer)
library(forecast)
library(ggplot2)
library(wooldridge)
# define time series
tsdat=ts(phillips, start=1948)

# Estimate AR(1)
res1 <- dynlm(unem ~ L(unem)      , data=tsdat, end=1996)
# AR(1) + lagged inflation
# practically this is the first equation of a VAR(1) system
res2 <- dynlm(unem ~ L(unem) + L(inf), data=tsdat, end=1996)
stargazer(res1, res2 ,type="text", keep.stat=c("n","adj.rsq","ser"))
```

Form the predictions and standard errors:  
```{r}
# AR(1) Predictions for 1997-2003 including 95% forecast intervals:
p1 <- predict(res1, newdata=window(tsdat,start=1997), interval="prediction")
p1
```

```{r}
# VAR(1) Predictions for 1997-2003 including 95% forecast intervals:
p2 <- predict(res2, newdata=window(tsdat,start=1997), interval="prediction")
p2
```

Plot the forecast from the AR(1) model
```{r message=FALSE, warning=FALSE}
unem <- ts(phillips$unem, start=1948)
AR1fit <- ts(p1, start=1997)
autoplot(unem) + autolayer(AR1fit) +geom_point(aes(y=unem)) + 
  geom_vline(xintercept = 1996, linetype=2) +
  ggtitle("Unemployment Forecasts for 1997-2003 using AR(1) Model")
```

Plot the forecast from the VAR(1) model
```{r} 
VAR1fit <- ts(p2, start=1997)
autoplot(unem) + 
  autolayer(VAR1fit) + 
  geom_vline(xintercept = 1996, linetype=2) +
  ggtitle("Unemployment Forecasts for 1997-2003 using VAR(1) Model")
```

Compute the out-of-sample forecasts and forecast errors:  
```{r}
# Actual unemployment and forecasts:
y  <- window(tsdat,start=1997)[,"unem"]
f1 <- predict( res1, newdata=window(tsdat,start=1997) )
f2 <- predict( res2, newdata=window(tsdat,start=1997) )

# Plot unemployment and forecasts:
matplot(time(y), cbind(y,f1,f2), type="l",  col="black",lwd=2,lty=1:3)
legend("topleft",c("Unempl.","Forecast 1","Forecast 2"),lwd=2,lty=1:3)
```


```{r}
# Forecast errors:
e1<- y - f1
e2<- y - f2
```


```{r}
# RMSE:
sqrt(mean(e1^2))
sqrt(mean(e2^2))
```


```{r}
# MAE:
mean(abs(e1))
mean(abs(e2))
```

The second model, VAR(1), a has smaller RMSE and MAE. 

# Fitting a VAR model

We can use `vars` package to estimate a VAR. 
For example, using the `phillips` data set: 
```{r}
library(fpp2)
library(vars)
tsdat <- ts(phillips, start=1948, end=1996)
var1 <- VAR(tsdat[,c("unem","inf")], p=1, type="const")
summary(var1)
```

Compute and plot forecasts: 

```{r}
forecast(var1) %>% 
  autoplot() +
  xlab("year")
```


The Granger-causality test from inflation to unemployment
```{r}
causality(var1, cause = "inf")
```

The F statistic is 19.87 with a very small p-value. Thus, we reject the null hypothesis. **Inflation Granger-causes unemployment**. This implies that inflation is useful in forecasting one-period-ahead unemployment rate. 

The Granger-causality test from unemployment to inflation: 
```{r}
causality(var1, cause = "unem")
```

The F test statistic is 0.06 with pvalue = 0.80, thus we don't reject the null hypothesis. **Unemployment does not Granger-cause inflation**. 

Overall, we can say that while inflation is useful in predicting the future unemployment rate the reverse is not true. Unemployment is not useful in predicting future inflation. 

<br>
<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>

 