---
title: "Lab 03: Introduction to Time Series Regressions - Part 1"
subtitle: "Econometrics II - YTU"
author: 
  name: "Prof. Dr. Hüseyin Taştan"
  affiliation: "Yıldız Technical University"
# date: "`r format(Sys.time(), '%d %B %Y')`"
date: 2021 Spring
output: 
  html_document:
    number_sections: true
    theme: lumen
    highlight: haddock 
    # code_folding: show
    toc: yes
    toc_depth: 3
    toc_float: yes
    keep_md: true
---
<style type="text/css"> 
body{
  background-color: #FAFAFA;
  font-size: 18px;
  line-height: 1.8;
}
code.r{
  font-size: 12pt;
}
</style>
<br>

```{r setup, include=FALSE}
# knitr::opts_chunk$set(echo = TRUE, results = 'asis', fig.show = 'asis')
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
``` 

# Replicating Examples in ch.10 of Wooldridge's Text (6th ed) 

## Example 10.1: Static Phillips Curve

```{r message=FALSE, warning=FALSE, results="hold"}
# Install packages
library(wooldridge)
library(dynlm)
library(lmtest)
library(car)
```

We first need to define a time series (ts) object in R. 
```{r}
# create a ts object 
inf <- ts(phillips$inf, start=1948, frequency = 1)
unem <- ts(phillips$unem, start=1948, frequency = 1)
# all variables in the data set: 
tsphillips <- ts(phillips, start=1948, frequency = 1)
tsphillips
```

### Time series plots

Using the base R plot function, we have 
```{r}
plot(unem)
plot(inf)
# ts.plot(tsphillips[,2:3])

ts.plot(unem,inf, col = c("black", "red"), lty=1:2, xlab="")
legend("topleft", bty="n", lty=1:2, col=c("black","red"),
       legend=c(" unem ", " inf "))
```

Altarnatively, we can use `ggplot()` as follows
```{r}
library(ggplot2)
ggplot(data = phillips, aes(year,inf)) + 
  geom_line() + 
  ylab("Inflation") + 
  xlab("Year")
```


Both inflation and unemployment on the same graph: 
```{r}
ggplot(data = phillips) +
  geom_line(aes(year, inf, color="inflation")) + 
  geom_line(aes(year, unem, color="unemployment")) +
  labs(color="") + 
  ylab("")
```

Here is another way which uses `forecast::autoplot()` function: 
```{r}
library(forecast)
library(ggplot2)
autoplot(inf, series = "inflation") + 
  autolayer(unem, series = "unemployment") +
  xlab("") + ylab("") + 
  ggtitle("Inflation and Unemployment Rate: 1948-2003")
```


### Static regression

```{r}
# using lm function
res <- lm(inf ~ unem, data = phillips)
summary(res)
```


```{r}
# or using dynlm package 
res1 <- dynlm(inf ~ unem, data  = tsphillips)
summary(res1)
```

The results suggest that inflation and unemployment move in the same direction, contradicting the Phillips curve theory (but we will see that there are several problems with regressions like this). 

## Example 10.2: Effects of Inflation and Deficits on Interest Rates

```{r}
# create a ts object 
i3 <- ts(intdef$i3, start=1948, frequency = 1)
inf <- ts(intdef$inf, start=1948, frequency = 1)
# plot
ts.plot(i3,inf, col = c("black", "red"), lty=1:2, xlab="")
legend("topleft", bty="n", lty=1:2, col=c("black","red"),
       legend=c(" i3 ", " inf "))
```

```{r}
ggplot(data = intdef) +  
  geom_line(aes(year, i3, color="i3")) + 
  geom_line(aes(year, inf, color="inf")) + 
  labs(color="") +
  ylab("") +
  theme_bw()
```

```{r}
tsintdata <- ts(intdef, start = 1948, frequency = 1)
i3_inf <- ts(tsintdata[,2:3], 
             start = 1948, frequency = 1)
head(i3_inf)
```

Using `ggfortify` package, plot inf and i3 together: 
```{r}
library(ggfortify)
autoplot(i3_inf)
```


```{r}
# static model:
intmodel1 <- lm(i3 ~ inf + def, data = intdef)
summary(intmodel1)
```

```{r}
# static model using dynlm package 
intmodel2 <- dynlm(i3 ~ inf + def, data  = tsintdata)
summary(intmodel2)
```


## Example 10.3: Puerto Rican Employment and the Minimum Wage

```{r}
minwage <- lm(lprepop ~ lmincov + lusgnp, data = prminwge)
summary(minwage)
```

## Example 10.4: Effects of Personal Exemption on Fertility Rates

```{r}
data(fertil3, package='wooldridge')
tsfertil <- ts(fertil3, start=1913)

# graphs
ts.plot(tsfertil[,1:2], col = c("black", "red"), lty=1:2, xlab="")
legend("topleft", bty="n", lty=1:2, col=c("black","red"),
       legend=c(" gfr ", " pr "))
```

```{r}
# Static model 
staticres <- dynlm(gfr ~ pe +  ww2 + pill, data = tsfertil)
summary(staticres)
```

```{r}
# Dynamic model using dynlm package
# Note: L(pe) is the first lag of pe, L(pe,2) is the second lag
dynres <- dynlm(gfr ~ pe + L(pe) + L(pe,2) + ww2 + pill, data = tsfertil)
summary(dynres)
coeftest(dynres)
```

```{r}
# F test. H0: all pe coefficients are=0
linearHypothesis(dynres, matchCoefs(dynres,"pe"))
# Calculating the LRP
b <- coef(dynres)
b["pe"]+b["L(pe)"]+b["L(pe, 2)"]
```

```{r}
# F test. H0: LRP=0
linearHypothesis(dynres,"pe + L(pe) + L(pe, 2) = 0")
```

 
## Example 10.5: Antidumping Filings and Chemical Imports

```{r}
impts <- ts(barium$chnimp, start=c(1978,2), frequency=12)

# plot time series
plot(impts)
```

```{r}
# Static model 
bariumres <- dynlm(log(chnimp) ~ log(chempi) + log(gas) + log(rtwex) + 
                     befile6 + affile6 + afdec6, data = barium)
summary(bariumres)
```

## Example 10.7: Housing Investment and Prices

```{r message=FALSE, warning=FALSE}
library(dynlm)
library(stargazer)
data(hseinv, package = 'wooldridge')

# Define Yearly time series beginning in 1947
tsdata <- ts(hseinv, start=1947)
```

```{r}
library(tsibble)
tbledata <- as_tsibble(tsdata, index=year, pivot_longer = FALSE)
# ts plot
ggplot(data = tbledata) +  
  geom_line(aes(year, log(invpc), color="log(invpc)")) + 
  geom_line(aes(year, log(price), color="log(price)")) +   
  labs(color="") +  
  ylab("") +
  theme_bw() +
  scale_color_brewer(palette = "Set1")
```



```{r, warning=FALSE}
# Linear regression of model with lags:
res1 <- dynlm(log(invpc) ~ log(price), data = tsdata)
res2 <- dynlm(log(invpc) ~ log(price) + trend(tsdata), data = tsdata)

# regression table
stargazer(res1,res2, type="text")
```


## Example 10.8: Fertility equation with linear and quadratic trend 


```{r}
# Static model with linear trend
fertilres1 <- dynlm(gfr ~ pe +  ww2 + pill + trend(tsfertil), data = tsfertil)
summary(fertilres1)
```


```{r}
# Static model with linear trend
data(fertil3, package='wooldridge')
fertil3$trend <- 1:nrow(fertil3)
fertil3$trendsq <- fertil3$trend^2
tsfertil <- ts(fertil3, start=1913)
fertilres2 <- lm(gfr ~ pe +  ww2 + pill + trend + trendsq, data = tsfertil)
summary(fertilres2)
```

## Example 10.9: Puerto Rican Employment Model with Trend

```{r}
# add a linear trend 
tsprminwage <- ts(prminwge, start=1950)
minwage <- dynlm(lprepop ~ lmincov + lusgnp + 
                   trend(tsprminwage, scale=FALSE), data = tsprminwage)
summary(minwage)
```

## Example 10.11: Effects of Antidumping Filings with Seasonality

```{r}
# add seasonal dummies
tsbarium <- ts(barium, start=c(1978,2), frequency = 12)
bariumres <- dynlm(log(chnimp) ~ log(chempi) + log(gas) + log(rtwex) + 
                                befile6 + affile6 + afdec6 + trend(tsbarium) + 
                                season(tsbarium), data = tsbarium)
summary(bariumres)
```


```{r}
# test joint significance of seasonal dummies 
# F test. H0: all seasonal coefficients are=0
linearHypothesis(bariumres, matchCoefs(bariumres,"season"))
```

P value is 0.59 so we do not reject the null hypothesis. The seasonal effects are jointly insignificant. 

# Example: Turkish Quarterly Real GDP 

```{r}
### TR RGDP seasonally adjusted
library(dplyr)
library(ggplot2)
library(fpp2)
load("../Data/RGDPTR.RData")
lrgdp <- log(rgdp)
autoplot(lrgdp) + 
  xlab("") +
  ggtitle("Log of Real GDP (Turkey): 1998q1-2019q3")
```

```{r}
# Linear trend fit
lintrend <- tslm(lrgdp ~ trend)
summary(lintrend)
```

**Exercise** Write down the linear trend output in equation form. What's the quarterly average growth rate? What's the annual average growth rate? 

Plot the residuals: 
```{r}
resid <- residuals(lintrend)
autoplot(resid) 
```

Forecasting using linear trend model: 
```{r}
library(forecast)
# Linear trend forecasts into 8 period ahead (2 years)
linfor <- forecast(lintrend, h = 8) 
autoplot(linfor) 
```


<br>
<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>


